SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-09-02 swmpr_manu.Rnw
created plos reject branch for future ref
working version following R journal wrapper file

% Template for PLoS
% Version 3.1 February 2015

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
\usepackage{setspace} 
\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

% for acronyms
\usepackage{acronym}

% packages I added
\usepackage{cleveref}
\usepackage[colorlinks=true,allcolors=Gray]{hyperref}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage[acronym]{glossaries}
\makeglossaries
\include{glossary}

%% Include all macros below

% cleveref options
\crefname{table}{Table}{Tables}
\crefname{figure}{Fig.}{Figs.}
\renewcommand{\figurename}{Fig.}

@@

% acronyms from glossaries package
\newacronym{CDMO}{CDMO}{Centralized Data Management Office}
\newacronym{CRAN}{CRAN}{Comprehensive R Archive Network}
\newacronym{DO}{DO}{dissolved oxygen}
\newacronym{NERRS}{NERRS}{National Estuarine Research Reserve System}
\newacronym{QAQC}{QAQC}{quality assurance/quality control}
\newacronym{SWMP}{SWMP}{System Wide Monitoring Program}

%% END MACROS SECTION

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.00\\textwidth')
options(replace.assign=TRUE,width=70,digits=1)
@

\begin{document}
\vspace*{0.35in}
\urlstyle{same}

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Marcus William Beck\textsuperscript{1, *}

\bigskip
\bf{1} ORISE Research Participation Program, USEPA National Health and Environmental Effects Research Laboratory, Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32651, USA
\\
\bigskip

* beck.marcus@epa.gov

\end{flushleft}

\section*{Abstract}
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making.  One example in the United States is the \gls{SWMP} that was implemented in 1995 by the federally-funded \acrlong{NERRS}.  This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries.  \gls{SWMP} data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with \gls{SWMP} environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
\linenumbers
\glsaddall
\printglossaries

\section*{Introduction}

@@

\subsection*{Data retrieval}

Two basic approaches using SWMPr are available to import \gls{SWMP} data into R, either through direct download or by importing local data (\Cref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using \gls{CDMO} web services. To do so, the IP address for the computer making the request must be registered by following instructions on the \gls{CDMO} website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with \gls{CDMO}.

@@

descrips <- c(
  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the \\gls{CDMO} using the zip downloads option.',
  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
  )

@@

# view all attributes of dat
atts <- paste0('\\texttt{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class.',
  'Row names of the data set, inherited from the \\texttt{data.frame} object class.',
  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}.',
  'Station identifier used by \\gls{NERRS} as a string with 7 or 8 characters.',
  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}, \\texttt{`turb'}, etc.",
  'Indicates if \\gls{QAQC} columns are present in the raw data.',
  'Start and end dates for the data.',
'Timezone of the station using the city/country format\\textsuperscript{a}.',
  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.'
)

@@

A sample dataset can be downloaded for use with the examples below (see the \nameref{supp_info}).  This dataset has an identical format as the data returned from the zip downloads feature of the \gls{CDMO}.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function, as shown above.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

\subsection*{Data organizing}

The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (\Cref{tab:organize}).

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified \gls{QAQC} flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding \gls{QAQC} column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the \gls{QAQC} column range from -5 to 5 to indicate the \gls{QAQC} flag that was assigned by \gls{CDMO} during initial processing.  The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} \gls{QAQC} flag (i.e., passed initial \gls{CDMO} checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the \gls{QAQC} columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

descrips <- c(
  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove \\gls{QAQC} columns and remove data based on \\gls{QAQC} flag values for a \\texttt{swmpr} object.  Only applies if \\gls{QAQC} columns are present. ', 
  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each \\gls{QAQC} flag used by \\gls{CDMO}.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \\texttt{comb}.',
  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
  )

@@

Viewing the number of observations for each parameter that are assigned to a \gls{QAQC} flag may be useful for deciding how to process the data with \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.

@@

Raw nutrient data obtained from the \gls{CDMO} will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if \gls{QAQC} columns are present.

@@

\subsection*{Data analysis}

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (\Cref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}). These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

@@

Time series can be smoothed to better characterize a signal from noisy data (\Cref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.  The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.  The \texttt{sides} argument specifies how the average is calculated for each observation.  Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (\Cref{fig:interp_ex}).  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.  An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.  The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.

@@

The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.  The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (\Cref{fig:decomp_ex1}).  An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.  The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.  For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).  The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).  For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.  A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.  For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\texttt{ts}} function for details).  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.  The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.  For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  \Cref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

Detailed exploratory graphics are also useful for evaluating general trends in observed data.  Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (\Cref{fig:summary_ex}).  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.  An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all \gls{SWMP} sites (see the \nameref{swmp_apps} section).

@@

Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}. The diffusion-corrected \gls{DO} flux estimates are averaged separately over each day and night of the time series. The nighttime average \gls{DO} flux is used to estimate respiration rates, while the daytime \gls{DO} flux is used to estimate net primary production. To generate daily integrated rates, respiration rates are assumed constant such that hourly night time \gls{DO} flux rates are multiplied by 24. Similarly, the daytime \gls{DO} flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production. Respiration rates are subtracted from daily net production estimates to yield gross production rates. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the \gls{DO} time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

The following is an example that shows use of the \texttt{ecometab} function from a combined water quality and weather data set.  Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (\Cref{fig:metab_ex}).

@@

\subsection*{Miscellaneous functions}

Several additional functions are provided that do not fit the above categories (\Cref{tab:misc}).  These functions are generally used within the main functions but may be useful for more customized evaluation of \gls{SWMP} data.

@@

\section*{Applications using the SWMPr package}
\label{swmp_apps}

>>>>

\documentclass[a4paper]{report}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{RJournal}
\usepackage{amsmath,amssymb,array}
\usepackage{booktabs}

%% load any required packages

@@

\begin{document}

%% do not edit, for illustration only
\sectionhead{Contributed research article}
\volume{XX}
\volnumber{YY}
\year{20ZZ}
\month{AAAA}

\begin{article}

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!h', warning = F, background = 'white', highlight = FALSE, prompt = TRUE, size = 'small', out.width = '0.00\\textwidth')
options(replace.assign=TRUE,width=70,digits=1)
knit_hooks$set(document = function(x) {sub('\\usepackage[]{color}', '\\usepackage{xcolor}', x, fixed = TRUE)})
@

\title{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
\author{by Marcus W Beck}

\maketitle

\abstract{
Standardized monitoring programs have vastly improved the quantity and quality of data that form the basis of environmental decision-making. One example in the United States is the System-Wide Monitoring Program (SWMP) that was implemented in 1995 by the federally-funded National Estuarine Research Reserve System. This program has provided two decades of continuous monitoring data at over 140 fixed stations in 28 estuaries. SWMP data have been used in a variety of applications with the general objective of describing dynamics of estuarine ecosystems to better inform effective coastal management.  However, simple tools for processing and evaluating the increasing quantity of data provided by the monitoring network have prevented broad-scale comparisons between systems and, in some cases, simple trend analysis of water quality parameters at individual sites.  I describe SWMPr, an open-source R package, for use with SWMP environmental data.  The package provides several functions that facilitate data retrieval, organization, and analysis of time series data to describe water quality, weather, and nutrient dynamics in the reserve estuaries.  Previously unavailable functions for estuaries are also provided to estimate rates of ecosystem metabolism using the open-water method.  Tools included with the SWMPr package have facilitated a cross-reserve comparison of trends, including simple evaluation of changes over time and comparisons of patterns in primary productivity.  Overall, the package provides an effective approach to link quantitative information with analysis tools that will greatly inform management programs aimed at coastal protection and restoration.
}

\section{Introduction}

@@

\subsection{Data retrieval}

Two basic approaches using SWMPr are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using CDMO web services. To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.

@@

descrips <- c(
  'Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \\texttt{exportAllParamsXMLNew} function on web services.',
  'Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \\texttt{exportAllParamsDateRangeXMLNew}.', 
  'Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.',
  'Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \\texttt{exportSingleParamXMLNew} function on web services.', 
  'Metadata for all stations, wrapper to \\texttt{exportStationCodesXMLNew} function on web services.',
  'Metadata for all stations at a single site, wrapper  to \\texttt{NERRFilterStationCodesXMLNew} function on web services.'
  )

@@

# view all attributes of dat
atts <- paste0('\\texttt{', names(attributes(dat)), '}')
atts <- gsub('_', '\\_', atts, fixed = T)
atts_class <- c('character', 'integer', 'character', 'character', 'character', 'logical', 'POSIXct', 'character', 'character')
atts_desc <- c(
  'Column names of the entire data set, inherited from the \\texttt{data.frame} object class.',
  'Row names of the data set, inherited from the \\texttt{data.frame} object class.',
  'Class of the data object indicating \\texttt{swmpr} and \\texttt{data.frame}.',
  'Station identifier used by NERRS as a string with 7 or 8 characters.',
  "Character vector of column names for data parameters, e.g., \\texttt{`do\\_mgl'}, \\texttt{`turb'}, etc.",
  'Indicates if QAQC columns are present in the raw data.',
  'Start and end dates for the data.',
'Timezone of the station using the city/country format\\textsuperscript{a}.',
  'Class of the \\texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.'
)

@@

A sample dataset can be downloaded for use with the examples below.  This dataset has an identical format as the data returned from the zip downloads feature of the CDMO.  These data are also included with the package as binary data files (RData) that can be loaded using the \texttt{data} function, as shown above.  These include \texttt{swmpr} objects for four stations at Apalachicola Bay: apacpnut, apacpwq, apadbwq, and apaebmet.  Information for each file can be viewed in the help documentation (e.g., \texttt{?apacpnut}).

\subsection{Data organizing}

The organize functions are used to clean or prepare the imported data for analysis, including viewing and removal of QAQC flags, subsetting, combining replicate nutrient observations, creating a standardized time series, and combining data of different types (Table~\ref{tab:organize}).

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.  The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} QAQC flag (i.e., passed initial CDMO checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

@@

descrips <- c(
  'Combines \\texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.',
  'Remove QAQC columns and remove data based on QAQC flag values for a \\texttt{swmpr} object.  Only applies if QAQC columns are present. ', 
  'View a summary of the number of observations in a \\texttt{swmpr} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.', 
  'Remove replicate nutrient data that occur on the same day.  The default is to average replicates.', 
  'Format data from a \\texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \\texttt{comb}.',
  'Subset by dates and/or columns for a \\texttt{swmpr} object.  This is a method passed to the generic \\texttt{subset} function provided in the base package.'
  )

@@

Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.

@@

Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.

@@

\subsection{Data analysis}

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (Table~\ref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}). These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

@@

Time series can be smoothed to better characterize a signal from noisy data (Figure~\ref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.  The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.  The \texttt{sides} argument specifies how the average is calculated for each observation.  Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (Figure~\ref{fig:interp_ex}).  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.  An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.  The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.

@@

The disaggregation of time series into additive or multiplicative components that can be attributed to separate sources of variance is another common application for trend analysis.  The \texttt{decomp} function is a simple wrapper to \texttt{decompose} \cite{Kendall83} that separates a time series into components describing a trend, cyclical variation (e.g., daily or annual), and the remainder  (Figure~\ref{fig:decomp_ex1}).  An additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), otherwise a multiplicative decomposition can be used.  The \texttt{frequency} argument describes the periodicity of the cyclical parameter in units of the native time step.  For example, the \texttt{frequency} for a parameter with daily periodicity would be 96 if the time step is 15 minutes (24 hours * 60 minutes / 15 minutes).  The frequency of a parameter with annual periodicity at a 15 minute time step would be 35040 (365 days * 24 hours * 60 minutes / 15 minutes).  For simplicity, character strings of \texttt{`daily'} or \texttt{`annual'} can be supplied in place of numeric values, although any number can be used to identify an arbitrary cyclical component.  A starting value of the time series must be supplied in the latter case that indicates the sequence in the cycle for the first observation.  For example, the starting value would be 1 if the first observation is at sunrise for a diurnal cycle (see the help file for the \href{https://stat.ethz.ch/R-manual/R-devel/library/stats/html/ts.html}{\texttt{ts}} function for details).  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

@@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package \cite{Cloern10,Jassby14}.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach works best for nutrient data that are typically obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data must first be aggregated on the monthly scale before decomposition.  The time series is decomposed into the grandmean, annual, seasonal, and events components, as compared to trend, seasonal, and random components for the \texttt{decomp} function described above.  For both, the random or events components can be considered anomalies that do not follow the trends in the remaining categories.  Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  Figure~\ref{fig:decomp_ex2} shows the results from the \texttt{decomp\_cj} function applied to a multi-year chlorophyll time series.

@@

Detailed exploratory graphics are also useful for evaluating general trends in observed data.  Several graphics showing seasonal and annual trends for a single SWMP parameter can be obtained using the \texttt{plot\_summary} function (Figure~\ref{fig:summary_ex}).  The plots include monthly distributions, monthly anomalies, and annual anomalies in multiple formats.  Anomalies are defined as the difference between the monthly or annual averages from the grand mean for the parameter.  An interactive web application \cite{Chang15} that uses this function is available for viewing results of any parameter at all SWMP sites (see the \nameref{swmp_apps} section).

@@

Finally, estimates of ecosystem metabolism provide a measure of overall system productivity to evaluate whether an ecosystem is a net source or sink of organic material.  The open-water method \cite{Odum56} is a common approach to quantify metabolism using a mass balance equation that describes the change in dissolved oxygen over time from the balance between photosynthetic and respiration processes, corrected using an empirically constrained air-sea gas diffusion model \cite{Ro06,Thebault08}. The diffusion-corrected dissolved oxygen (DO) flux estimates are averaged separately over each day and night of the time series. The nighttime average DO flux is used to estimate respiration rates, while the daytime DO flux is used to estimate net primary production. To generate daily integrated rates, respiration rates are assumed constant such that hourly night time DO flux rates are multiplied by 24. Similarly, the daytime DO flux rates are multiplied by the number of daylight hours, which varies with location and time of year, to yield net daytime primary production. Respiration rates are subtracted from daily net production estimates to yield gross production rates. The metabolic day is considered the 24 hour period between sunsets on two adjacent calendar days

The \texttt{ecometab} function is used to implement the open-water method with a combined water quality and weather dataset \cite{Caffrey14}.  Several assumptions must be met for a valid interpretation of the results.  First, the DO time series is assumed to be a sample of the same water mass over time.  Tidal advection may have a significant influence on the time series, which can contribute to a substantial amount of noise in metabolic estimates.  The extent to which tidal advection influences the dissolved oxygen signal depends on various site-level characteristics and an intimate knowledge of the site may be required.  Second, areal rates for gross production and total respiration are based on volumetric rates normalized to the depth of the water column at the sampling location, which is assumed to be well-mixed, such that the water quality sensor is reflecting the integrated processes in the entire water column (including the benthos).  Water column depth is calculated as the mean value of the depth variable across the time series in the \texttt{swmpr} object. Depth values are floored at one meter for very shallow stations and 0.5 meters is also added to reflect the practice of placing sensors slightly off of the bottom.  Third, the air-sea gas exchange model is calibrated with wind data either collected at, or adjusted to, wind speed at 10 m above the surface.  The metadata should be consulted for exact height. Other assumptions may apply and relevant resources should be consulted to ensure appropriate application of the open-water method (see \cite{Kemp12,Needoba12}).

The following is an example that shows use of the \texttt{ecometab} function from a combined water quality and weather data set.  Monthly aggregations of the raw, daily estimates are plotted using \texttt{plot\_metab} (Figure~\ref{fig:metab_ex}).

@@

\subsection{Miscellaneous functions}

Several additional functions are provided that do not fit the above categories (Table~\ref{tab:misc}).  These functions are generally used within the main functions but may be useful for more customized evaluation of SWMP data.

@@

\section{Applications using the SWMPr package}
\label{swmp_apps}
